Cooling Requirements for the Vertical Shear Instability, \\ with Implications for Protoplanetary Disks

### 2015-04-08 paper.tex
minor tweaks to the early part of intro.  the intro is quite good already.  my current concern is that we should not ONLY give credit to BL15 for the physical interpretation of VSI/GSFI, as this potentially gives the false impression that it was not understood before.  I'll tweak that a bit next.

However, MHD turbulence may not be applicable to all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent MHD simulations
suggest significant portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. Thus, alternative mechanisms 
to angular momentum transport in PPDs, such as purely hydrodynamic
turbulence,  remain of interest \citep{armitage10}. In addition to
mass accretion, hydrodynamic turbulence in PPDs will also
impact dust evolution and planetesimal formation \citep{chiang10}.

One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI). The VSI is an 
axisymmetric linear instability 
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height $z$ from the disk midplane. This
dependence occurs when the equilibrium pressure depends on both the
temperature and density, making the disk baroclinic. It is therefore
generally expected that $\p_z\Omega\neq 0$ in astrophysical disks.

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
A height-dependence in $\Omega$, or vertical shear, can render
inertial waves unstable. \cite{barker15} gives a simple 
interpretation of the VSI: in an axisymmetric, inviscid, neutrally
stratified disk, the exchange of fluid parcels vertically, while
conserving their specific angular momentum, can result in a new
configuration of lower energy if the distribution of specific angular
momentum varies with $z$. The VSI is characterized by perturbations
with large vertical length-scales  relative to radial length-scales.

>>>>

However, MHD turbulence may not be applicable to all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent MHD simulations
suggest significant portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. Thus, alternative mechanisms 
to angular momentum transport in PPDs, such as purely hydrodynamic
turbulence,  remain of interest \citep{armitage10}. In addition to
mass accretion, turbulence in PPDs will also
impact dust evolution and planetesimal formation \citep{chiang10}.

One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI). The VSI is an 
axisymmetric linear instability 
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height $z$ from the disk midplane. 
Vertical shear occurs whenever the disk is baroclinic, i.e.\ when constant 
density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
practically unavoidable in disks, except at special locations like the midplane.

The VSI  in disks is fundamentally equivalent to the
Goldreich-Schubert-Fricke instability 
\citep[GSFI,][]{goldreich67,fricke68} in differentially rotating stars.   
A height-dependence in $\Omega$, or vertical shear, can render
inertial waves unstable. \cite{barker15} gives a simple 
interpretation of the VSI: in an axisymmetric, inviscid, neutrally
stratified disk, the exchange of fluid parcels vertically, while
conserving their specific angular momentum, can result in a new
configuration of lower energy if the distribution of specific angular
momentum varies with $z$. The VSI is characterized by perturbations
with large vertical length-scales  relative to radial length-scales.
